---
title: "Cardiac maturation process from human iPSCs"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Introduction

#Summary

To improve the maturity of iPSC-cardiomyocytes, we addressed the cell type specification of iPSC-cardiomyocytes by microarray and analyzed in detail about cardiac maturation process.
We compared gene expression between samples, and extracted several key genes for cardac maturation and check the function of these genes. We show that a number of genes expression correlate with cardiac maturation process, and found surface markers to discriminate differentiated cardiomyocytes from other immature cells.

#Overall design

mRNA profiles of differentiated iPSC-cardiomyocytes at several differentiation stages. Additionally, we checked the function of several important genes for cardiac maturation, which were extracted from microarray data.


```{r}
iPS <- read.csv("C:/Users/usuari/Desktop/bioinfo/omics/iPS.txt", sep="")
```

```{r}
expressions <- read.delim("C:/Users/usuari/Desktop/bioinfo/omics/GSE72611_series_matrix_cut.txt", header=TRUE,row.names = 1)
```

```{r}
dim(expressions)
summary(expressions)
```

```{r}
boxplot(expressions, col = c(2,2,2,2,3,3,3,3,4,4,4,4,5,5,5,5,5))
```

