---
title: "exercise 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This is a Microarray Data Analysis

```{r}
#FOLDER DESTINATION DEFINITIONS
workingDir <- getwd()
workingDir
dataDir <- file.path(workingDir, "dades")
dataDir
resultsDir <- file.path(workingDir, "results")
resultsDir
setwd(resultsDir)
```

We install all the packages we need
```{r}
##INSTALLATION OF PACKAGES NEEDED

if (!require(BiocManager)) install.packages("BiocManager")

installifnot <- function (pkg){
  if (!require(pkg, character.only=T)){
    BiocManager::install(pkg)
  }else{
    require(pkg, character.only=T)
  }
}

installifnot("pd.mogene.1.0.st.v1")
installifnot("mogene10sttranscriptcluster.db")
installifnot("oligo")
installifnot("limma")
installifnot("Biobase")
installifnot("arrayQualityMetrics")
installifnot("genefilter")
installifnot("multtest")
installifnot("annotate")
installifnot("xtable")
installifnot("gplots")
installifnot("scatterplot3d")
```

we load all the necessary data. In this case we load the targets and the Expression Set 
```{r}
##LOAD DATA: TARGETS AND EXPRESSION SET. 

#TARGETS
targets <- read.csv(file=file.path(dataDir,"targets.txt"), sep="")
targets$color <- c(0,2,2,3,3,3,4,4,5,5,5,5,5,5,6,6,6)
targets$ShortName <- c('UD67','D68','D69','CM70','CM71','CM72','Chigh73','Clow74','HDAC75','HDAC76','HDAC77','HDAC78','HDAC79','HDAC80','NC81','NC82','NC83')
targets$group <- c('None','None','None','None','None','None','CORIN','CORIN','HDAC','HDAC','HDAC','HDAC','HDAC','HDAC','NC','NC','NC')
targets
```

```{r}
#RAWDATA
if (!require(GEOquery)) {
  BiocManager::install("GEOquery")
}
require(GEOquery)
gse <- getGEO("GSE72611")
class(gse)
names(gse)
gse[[1]]
esetFromGEO <- gse[[1]]
rawData <- esetFromGEO

#DEFINE SOME VARIABLES FOR PLOTS
sampleNames <- as.character(targets$ShortName)
sampleColor <- as.character(targets$color)
```

We use this raw data to create a boxplot
```{r}
##QUALITY CONTROL OF ARRAYS: RAW DATA

#BOXPLOT
boxplot(rawData, which="all",las=2, main="Intensity distribution of RAW data", 
        cex.axis=0.6, col=sampleColor, names=sampleNames)
```

We do a hierarquical clustering of the raw data
```{r}
#HIERARQUICAL CLUSTERING
clust.euclid.average <- hclust(dist(t(exprs(rawData))),method="average")
plot(clust.euclid.average, labels=sampleNames, main="Hierarchical clustering of RawData", 
     cex=0.7,  hang=-1)
```

We do a PCA plot to see the distribution of this raw data
```{r}
#PRINCIPAL COMPONENT ANALYSIS
plotPCA <- function ( X, labels=NULL, colors=NULL, dataDesc="", scale=FALSE, formapunts=NULL, myCex=0.8,...)
{
  pcX<-prcomp(t(X), scale=scale) # o prcomp(t(X))
  loads<- round(pcX$sdev^2/sum(pcX$sdev^2)*100,1)
  xlab<-c(paste("PC1",loads[1],"%"))
  ylab<-c(paste("PC2",loads[2],"%"))
  if (is.null(colors)) colors=1
  plot(pcX$x[,1:2],xlab=xlab,ylab=ylab, col=colors, pch=formapunts, 
       xlim=c(min(pcX$x[,1])-100000, max(pcX$x[,1])+100000),ylim=c(min(pcX$x[,2])-100000, max(pcX$x[,2])+100000))
  text(pcX$x[,1],pcX$x[,2], labels, pos=3, cex=myCex)
  title(paste("Plot of first 2 PCs for expressions in", dataDesc, sep=" "), cex=0.8)
}

plotPCA(exprs(rawData), labels=sampleNames, dataDesc="raw data", colors=sampleColor,
        formapunts=c(rep(16,4),rep(17,4)), myCex=0.6)
```

We can see that all the data is clustered in one point

We save all these plots into a file
```{r}
#SAVE TO A FILE
pdf(file.path(resultsDir, "QCPlots_Raw.pdf"))

boxplot(rawData, which="all",las=2, main="Intensity distribution of RAW data", 
        cex.axis=0.6, col=sampleColor, names=sampleNames)
plot(clust.euclid.average, labels=sampleNames, main="Hierarchical clustering of samples of RawData", 
     cex=0.7,  hang=-1)
plotPCA(exprs(rawData), labels=sampleNames, dataDesc="raw data", colors=sampleColor,
        formapunts=c(rep(16,4),rep(17,4)), myCex=0.6)
dev.off()
```

We normalize the data
```{r}
##DATA NORMALIZATION
eset <- rawData

write.exprs(eset, file.path(resultsDir, "NormData.txt"))

```

We redo the plots with the normalized data. 
```{r}

##QUALITY CONTROL OF ARRAYS: NORMALIZED DATA


#BOXPLOT
boxplot(eset, las=2, main="Intensity distribution of Normalized data", cex.axis=0.6, 
        col=sampleColor, names=sampleNames)

#HIERARQUICAL CLUSTERING
clust.euclid.average <- hclust(dist(t(exprs(eset))),method="average")
plot(clust.euclid.average, labels=sampleNames, main="Hierarchical clustering of NormData", 
     cex=0.7,  hang=-1)

#PRINCIPAL COMPONENT ANALYSIS
plotPCA <- function ( X, labels=NULL, colors=NULL, dataDesc="", scale=FALSE, formapunts=NULL, myCex=0.8,...)
{
  pcX<-prcomp(t(X), scale=scale) # o prcomp(t(X))
  loads<- round(pcX$sdev^2/sum(pcX$sdev^2)*100,1)
  xlab<-c(paste("PC1",loads[1],"%"))
  ylab<-c(paste("PC2",loads[2],"%"))
  if (is.null(colors)) colors=1
  plot(pcX$x[,1:2],xlab=xlab,ylab=ylab, col=colors, pch=formapunts, 
       xlim=c(min(pcX$x[,1])-10, max(pcX$x[,1])+10),ylim=c(min(pcX$x[,2])-10, max(pcX$x[,2])+10))
  text(pcX$x[,1],pcX$x[,2], labels, pos=3, cex=myCex)
  title(paste("Plot of first 2 PCs for expressions in", dataDesc, sep=" "), cex=0.8)
}

plotPCA(exprs(eset), labels=sampleNames, dataDesc="NormData", colors=sampleColor,
        formapunts=c(rep(16,4),rep(17,4)), myCex=0.6)

#SAVE TO A FILE
pdf(file.path(resultsDir, "QCPlots_Norm.pdf"))

boxplot(eset, las=2, main="Intensity distribution of Normalized data", cex.axis=0.6, 
        col=sampleColor, names=sampleNames)

plot(clust.euclid.average, labels=sampleNames, main="Hierarchical clustering of NormData", 
     cex=0.7,  hang=-1)

plotPCA(exprs(eset), labels=sampleNames, dataDesc="selected samples", colors=sampleColor,
        formapunts=c(rep(16,4),rep(17,4)), myCex=0.6)
dev.off()

#ARRAY QUALITY METRICS
arrayQualityMetrics(eset,  reporttitle="QualityControl", force=TRUE)
```

We can see the difference in the PCA plot when the data is normalized

Now we filter the data
```{r}

##FILTER OUT THE DATA

annotation(eset) <- "org.Mm.eg.db"
eset_filtered <- nsFilter(eset, var.func=IQR,
                          var.cutoff=0.75, var.filter=TRUE,
                          filterByQuantile=TRUE)
write.csv2(eset_filtered$eset,file.path(resultsDir, "Filtered.csv2"))
#NUMBER OF GENES OUT
print(eset_filtered$filter.log$numLowVar)

#NUMBER OF GENES IN
print(eset_filtered$eset)
```

We use this normalized and filtered data to do the comparisons between groups
```{r}

##DIFERENTIAL EXPRESSED GENES SELECTION. LINEAR MODELS. COMPARITIONS


#CONTRAST MATRIX.lINEAR MODEL
treat <- targets$group
lev <- factor(treat, levels = unique(treat))
design <-model.matrix(~0+lev)
colnames(design) <- levels(lev)
rownames(design) <- sampleNames
print(design)

#COMPARISON
cont.matrix1 <- makeContrasts( 
  None.vs.CORIN = None-CORIN,None.vs.HDAC = None-HDAC,None.vs.NC = None-NC,
  levels = design)
comparison1 <- "None.vs.CORIN"
comparison2 <- "None.vs.HDAC"
comparison3 <- "None.vs.NC"
cont.matrix1
#MODEL FIT
fit1 <- lmFit(eset_filtered$eset, design)
fit.main1 <- contrasts.fit(fit1, cont.matrix1)
fit.main1 <- eBayes(fit.main1)

```

We create data frames of each comparison
```{r}

##DIFERENTIAL EXPRESSED GENES LISTS.TOPTABLES


#FILTER BY FALSE DISCOVERY RATE AND FOLD CHANGE
topTab_NvC <-  topTable (fit.main1, number=nrow(fit.main1), coef= "None.vs.CORIN", adjust="fdr")
head(topTab_NvC)

topTab_NvH <-  topTable (fit.main1, number=nrow(fit.main1), coef= "None.vs.HDAC", adjust="fdr")
head(topTab_NvH)

topTab_NvNC <-  topTable (fit.main1, number=nrow(fit.main1), coef= "None.vs.NC", adjust="fdr")
head(topTab_NvNC)


```

We use these data frames to create volcano plots and we save them in a file
```{r}

##VOLCANO PLOTS

volcanoplot(fit.main1, coef="None.vs.CORIN", highlight=10)
volcanoplot(fit.main1, coef="None.vs.HDAC", highlight=10)
volcanoplot(fit.main1, coef="None.vs.NC", highlight=10)

volcanoplot(fit.main1, highlight=10, names=fit.main1$ID)
abline(v = c(-3, 3))


pdf(file.path(resultsDir,"Volcanos.pdf"))
volcanoplot(fit.main1, highlight = 10, names = fit.main1$ID)
abline(v = c(-3, 3))
dev.off()
```

